# Surveys {#surveys}

I used an anonymous survey to ask colleagues some questions about which topics may be more interesting or useful in their research. 

## March 2022

### Question 1

In the first question people indicated topics of interests. The winner is multilevle models, followed closely by Bayesian statistics.

```{r, message=FALSE, echo=FALSE, warning=F,fig.height=5,fig.width=7,fig.align='center'}
library(tidyverse)

# load Qualtrics csv output (edited to remove double header and IP addresses)
d <- read_csv('../data/survey_march22/stats-survey-rhul_edit.csv')

d %>%
  filter(Finished==TRUE) %>%
  select(starts_with('models_') & !contains('TEXT')) %>%
  mutate(id = 1:nrow(.)) %>%
  pivot_longer(starts_with('models_'),names_to = 'what') %>%
  filter(!is.na(value)) %>%
  filter(value!="Other (please specify)") %>%
  mutate(value_short = str_extract(value, pattern='^[^:\\.)]*')) %>% 
  # simplify labels (could be more elegant if I knew regex better)
  mutate(value_short = ifelse(str_detect(value_short,"\\((?=e)"),
                              str_sub(value_short,1,-1 -2),
                              value_short)) %>%
  mutate(value_short = ifelse(str_detect(value_short,"\\("),
                              str_c(value_short,")"),
                              value_short)) %>%
  group_by(value_short) %>%
  summarise(N = length(id)) %>%
  arrange(N) %>%
  mutate(value_short=as_factor(value_short)) %>%
  ggplot(aes(y=value_short, x=N))+
  theme_gray(12) +
  geom_col() +
  labs(y="")
```

Additionally there were some additional suggestions.

```{r echo=FALSE}
d %>%
  filter(Finished==TRUE) %>%
  select(starts_with('models_') & contains('TEXT')) %>%
  filter(!is.na(models_17_TEXT)) %>%
  pull(models_17_TEXT)

```

A few other topics were mentioned in the comment section:

- Shiny apps
- Network modelling
- Longitudinal analyses
- Random forests
- Neural network


### Question 2

Here people indicated their interest for topics related to data analysis.

```{r, message=FALSE, echo=FALSE, warning=F,fig.height=4,fig.width=8,fig.align='center'}
d %>%
  filter(Finished==TRUE) %>%
  select(starts_with('extra_') & !contains('TEXT')) %>%
  mutate(id = 1:nrow(.)) %>%
  pivot_longer(starts_with('extra_'),names_to = 'what') %>%
  filter(!is.na(value)) %>%
  filter(value!="Other (please specify)") %>%
  group_by(value) %>%
  summarise(N = length(value)) %>%
  arrange(N) %>%
  mutate(value=as_factor(value)) %>%
  ggplot(aes(y=value, x=N))+
  theme_gray(12) +
  geom_col() +
  labs(y="")
```

Other things mentioned in the comments were:

- SPM
- Docker
- Python


### Question 4

This question was about likelihood of using different formats of support

```{r, message=FALSE, echo=FALSE, warning=F,fig.height=6,fig.width=6,fig.align='center'}
d %>%
  filter(Finished==TRUE) %>%
  select(starts_with('format_')) %>%
  mutate(id = 1:nrow(.)) %>%
  pivot_longer(starts_with('format_'),names_to = 'what') %>%
  mutate(value=factor(value, ordered=T, levels=c("Very unlikely",
                                                 "Unlikely",
                                                 "Unsure",
                                                 "Likely",
                                                 "Very likely"))) %>%
  filter(!is.na(value)) %>%
  mutate(what=case_when(
    str_detect(what,"1") ~ "Short workshop",
    str_detect(what,"2") ~ "Longer workshop",
    str_detect(what,"3") ~ "one-on-one meeting",
    str_detect(what,"4") ~ "shared resource",
    str_detect(what,"5") ~ "collaboration"
  ) ) %>%
  group_by(what, value) %>%
  summarise(N = length(id)) %>%
  ggplot(aes(x=value, y=N)) +
  geom_col()+
  facet_wrap(.~what, ncol=2) +
  labs(x="")
```

### Respondents' status

The final questions asked about the status / career level.

```{r, message=FALSE, echo=FALSE, warning=F,fig.height=4,fig.width=4,fig.align='center'}
d %>%
  filter(Finished==TRUE) %>%
  filter(!is.na(role)) %>%
  ggplot(aes(x=role))+
  geom_histogram(stat='count')
```


