# Workshops  {#workshops}

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r style-share-again, echo=FALSE}
xaringanExtra::style_share_again(
  share_buttons = c("twitter", "linkedin", "pocket")
)
```

Either click on the links or click on the embedded slides and press 'F' to go full screen. Advance through slides with arrow keys, and press 'O' for an overview of all slides.


## Linear multilevel models (LMM) workshop (9th Sept 2022)



**Part 1** ([link](https://mlisi.xyz/files/workshops/LMM101/LMM_part1.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/LMM101/LMM_part1.html",
  ratio = "16:9"
)

```

**Part 2** ([link](https://mlisi.xyz/files/workshops/LMM101/LMM_part2.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/LMM101/LMM_part2.html",
  ratio = "16:9"
)
```

**Practicals**

[Exercises with solution (link)](https://mlisi.xyz/files/workshops/LMM101/exercises/exercises_questions.html)


## Introduction to linear algebra

Lecture for MSc module PS5210 (7th December 2022)

**Part 2** ([link](https://mlisi.xyz/files/workshops/numerical_skills_lecture/slides_numerical_skills_part2.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/numerical_skills_lecture/slides_numerical_skills_part2.html",
  ratio = "16:9"
)
```


## Introduction to meta-analyses in R

22nd February 2023

([link](https://mlisi.xyz/files/workshops/meta_analyses/meta_analyses_slides.html))


```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/meta_analyses/meta_analyses_slides.html",
  ratio = "16:9"
)
```
