--- 
always_allow_html: true
---

# Workshops  {#workshops}

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r style-share-again, echo=FALSE}
xaringanExtra::style_share_again(
  share_buttons = c("twitter", "linkedin", "pocket")
)
```

Either click on the links or click on the embedded slides and press 'F' to go full screen. Advance through slides with arrow keys, and press 'O' for an overview of all slides.


## Linear multilevel models (LMM) workshop (9th Sept 2022)



**Part 1** ([link](https://mlisi.xyz/files/workshops/LMM101/LMM_part1.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/LMM101/LMM_part1.html",
  ratio = "16:9"
)

```

**Part 2** ([link](https://mlisi.xyz/files/workshops/LMM101/LMM_part2.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/LMM101/LMM_part2.html",
  ratio = "16:9"
)
```

**Practicals**

[Exercises with solution (link)](https://mlisi.xyz/files/workshops/LMM101/exercises/exercises_questions.html)



## Introduction to meta-analyses in R

22nd February 2023

([link](https://mlisi.xyz/files/workshops/meta_analyses/meta_analyses_slides.html))


```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/meta_analyses/meta_analyses_slides.html",
  ratio = "16:9"
)
```


## Power analyses via data simulation

15th November 2023

Links: ([slides](https://mlisi.xyz/files/workshops/power_analyses/power_analyses.html)), 
([script](https://mlisi.xyz/files/workshops/power_analyses/power_analysis_workshop_script.R))


```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/power_analyses/power_analyses.html",
  ratio = "16:9"
)
```



## Introduction to Bayesian data analysis using R and Stan

29th November 2023

Link: ([slides](https://mlisi.xyz/files/workshops/intro_bayes/intro-Bayes.html)), 

See this Github repository for code examples and datasets: [https://github.com/mattelisi/intro-Bayes](https://github.com/mattelisi/intro-Bayes).

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/intro_bayes/intro-Bayes.html",
  ratio = "16:9"
)
```



## Introduction to causal reasoning with graphical models

15th January 2025

Link: ([slides](https://mlisi.xyz/files/workshops/causal/RHUL_causal_intro.html)), 

<div>

```{=html}
  <iframe class="slide-deck" src="https://mlisi.xyz/files/workshops/causal/RHUL_causal_intro.html" style="width: 100%;"></iframe>
```
</div>





## Introduction to linear algebra

Lecture for MSc module PS5210 (version from: 7th December 2022)

**Part 2** ([link](https://mlisi.xyz/files/workshops/numerical_skills_lecture/slides_numerical_skills_part2.html))

```{r, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "https://mlisi.xyz/files/workshops/numerical_skills_lecture/slides_numerical_skills_part2.html",
  ratio = "16:9"
)
```

